---
title: "GAMs and Community Construction"
output: html_document
date: "2025-07-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Extracting relevant data

This code serves to clean up occurence data while retaining site location, CWD, and Temperature. Data sourced from Krishnadas et al., 2021.

```{r}
 
raw_data <- read.csv("/Users/aayushmanchalwar/Desktop/thesis/gams/data/geb13350-sup-0002-tree-species-occ.csv") 
raw_longitudes <- raw_data$Latitude
raw_latitudes <- raw_data$Longitude
raw_temp <- raw_data$tdryq
raw_CWD <- raw_data$CWD
raw_occ <- raw_data$sp.occ
raw_site <- raw_data$Plot_ID
raw_spname <- raw_data$Species

relevant_data <- data.frame(unlist(raw_longitudes), unlist(raw_latitudes), unlist(raw_CWD), unlist(raw_occ), unlist(raw_site), unlist(raw_spname), unlist(raw_temp))
colnames(relevant_data) <- c("Longitude", "Latitude", "CWD", "Occurence", "Site", "Species", "Temperature")

sitebyspecies <- data.frame(matrix(data = "", nrow = 286, ncol = 183))
rownames(sitebyspecies) <- c(unlist(unique(relevant_data$Site)))
colnames(sitebyspecies) <- c(unlist(unique(relevant_data$Species)))

```
`

## Visualising the study site

The following plots the CWD landscape of the Western Ghats with the sampling locations overlaid on it. There are a total of 286 sampling locations.

```{r}
library(ncdf4)
library(terra)
library("readxl")
library("CFtime")
library(lattice)
library(RColorBrewer)

OccLatLon <- data.frame(unlist(relevant_data$Latitude), unlist(relevant_data$Longitude))
names(OccLatLon) <- c("lat", "lon")
vect(OccLatLon, geom=c("lat", "lon"), crs="", keepgeom=FALSE)
CWDraster <- rast('/Users/aayushmanchalwar/Desktop/thesis/gams/data/CWD.nc 2')
ghats <- c(73,78,8,20)
WGraster<- crop(CWDraster,ghats)
plot(WGraster, col=map.pal("ryb"), background = "white")
points(OccLatLon, col = "black", pch = 1, cex = 0.3)

```


## Using GAMs

Generalized Additive Models are used to generate probability-of-occurence curves across the CWD gradient, given occurence data and on-site CWD values.

``` {r}
library("mgcv")

plotter <- function(species) #function to obtain the GAM object for an input species
  {
    indices <- which(relevant_data$Species == toString(species))
    testcwd <- vector(length = length(indices))
    testocc <- vector(length = length(indices))
    testtemp <- vector(length = length(indices))
    testlon <- vector(length = length(indices))
    testlat <- vector(length = length(indices))
    
    for (i in 1:length(indices))
    {
      testcwd[i] <- relevant_data$CWD[indices[i]]
      testocc[i] <- relevant_data$Occurence[indices[i]]
      testtemp[i] <- relevant_data$Temperature[indices[i]]
      testlon[i] <- relevant_data$Longitude[indices[i]]
      testlat[i] <- relevant_data$Latitude[indices[i]]
    }
    testdata <- data.frame(unlist(testcwd), unlist(testocc), unlist(testtemp), unlist(testlon), unlist(testlat))
    
    colnames(testdata) <- c("CWD", "Occurence", "Temperature", "Longitude", "Latitude")
    
    
    test <- gam(Occurence ~ s(CWD),data = testdata, family = "binomial"(link = "logit"), method = "REML")
    
    #plot(test, trans = plogis, pages = 1)
    return(test)
}

```



We now have everything needed to quantify response diversity. To do this, we use a divergence-based measure that represents the stabilising property of RD -- at each value of the climate variable (CWD in this case), a measure of the difference between the largest and smallest first derivatives of the performance curves is used to represent RD. 

Here, a divergence value of 1 represents an "ideal" community, corresponding to a situation where the derivatives symmetrically span zero. This in turn implies the existence of equally positive and negative environment responses in the community, and such an assemblage is likely to be stable.


```{r}
library(gratia)
library(dplyr)
library(tidyr)
library(ggplot2)

all_species <- data.frame(unique(relevant_data$Species))
all_species <- all_species[,1]
cwds <- relevant_data %>% group_by(Site) %>% summarize(cwd = mean(CWD))




full_preds <- data.frame(matrix(data = "", nrow = 183, ncol = 286)) #matrix of occurence probabilities at each CWD value/site
rownames(full_preds) <- all_species
colnames(full_preds) <- c(unlist(cwds$Site))

full_lowerCI <- data.frame(matrix(data = "", nrow = 183, ncol = 286)) #matrix of lower confidence interval for species occurence probabilities
rownames(full_lowerCI) <- all_species
colnames(full_lowerCI) <- c(unlist(cwds$Site))

for (i in 1:183){
  species <- rownames(full_preds)[i]
  sp_gam <- plotter(species)
  sp_predicted_prob <- predict.gam(sp_gam, se.fit = TRUE)
  
  full_preds[i,] <- sp_predicted_prob$fit
  full_lowerCI[i,] <- plogis(sp_predicted_prob$fit - 1.96*sp_predicted_prob$se.fit)
}

full_derivatives <- data.frame(matrix(data = "", nrow = 183, ncol = 286))
rownames(full_derivatives) <- all_species
colnames(full_derivatives) <- c(unlist(cwds$Site))  

for (i in 1:183){
  focal_species <- all_species[i]
  focal_gam <- plotter(focal_species)
  focal_derivative <- derivatives(focal_gam, data = data.frame(CWD = c(cwds$cwd)))
  full_derivatives[i,] <- focal_derivative$.derivative
}


empty <- data.frame(matrix(data = "", nrow = 286, ncol = 183))
rownames(empty) <- c(unlist(cwds[,1]))

threshold <- 0.05#if the lower CI value for occurrence probability of a species at a CWD value is above the threshold, then said species is deemed to exist there. 

for(i in 1:286){
  
  focal_site <- rownames(cwds)[i]
  indices <- which(full_lowerCI[,i] > threshold)
  species_at_site <- c(rownames(full_lowerCI)[c(indices)])
  row <- c(species_at_site, rep(NA, 183 - length(species_at_site)))
  empty[i,] <- row
  
}

full <- empty #this is a matrix of the species said to exist at each site

rd_at_site <- data.frame(matrix(data = "", nrow = 0, ncol = 3))


for (i in 1:286){
  focal_comm <- full[i,][!is.na(full[i,])]
  focal_site <- rownames(full)[i]
  focal_cwd <- cwds$cwd[cwds$Site == focal_site]
  slice <- full_derivatives[c(which(rownames(full_derivatives) %in% focal_comm)),which(colnames(full_derivatives) == focal_site)]
  slice <- as.numeric(slice)
  h <- max(slice)
  l <- min(slice)
  divergence <- ((h-l) - abs(abs(h) - abs(l)))/(h-l) #this is divergence
  #divergence <- abs(mean(slice)) # this is imbalance
  rd_at_site <- rbind(rd_at_site, c(focal_site, focal_cwd, divergence))
  
}

colnames(rd_at_site) <- c("Site", "CWD","RD")
rd_at_site[,2] <- as.numeric(rd_at_site[,2])
rd_at_site[,3] <- as.numeric(rd_at_site[,3])


full_gam <- gam(RD ~ s(CWD), family = "betar"(link = "logit"), data = rd_at_site) #gam for the RD-CWD relationship

new_cwd <- data.frame(CWD = seq(min(rd_at_site$CWD), max(rd_at_site$CWD), length.out = 200))



pred <- predict(full_gam, newdata = new_cwd, se.fit = TRUE)

new_cwd$fit <- plogis(pred$fit)
new_cwd$lower <- plogis(pred$fit - 2 * pred$se.fit)
new_cwd$upper <- plogis(pred$fit + 2 * pred$se.fit)


#plot
ggplot() +
  geom_point(data = rd_at_site, aes(x = CWD, y = RD), alpha = 0.5) +
  geom_ribbon(data = new_cwd, aes(x = CWD, ymin = lower, ymax = upper), alpha = 0.2, fill = "blue") +
  geom_line(data = new_cwd, aes(x = CWD, y = fit), color = "blue", linewidth = 1) +
  labs(title = "RD vs CWD",
       x = "CWD", y = "RD") +
  theme_minimal()



```






```{r}
#visualising RD across space

lats <- relevant_data %>% group_by(Site) %>% summarise(Lat = mean(Latitude))
lons <- relevant_data %>% group_by(Site) %>% summarise(Lon = mean(Longitude))
rd_at_site$Lat <- unlist(lats[,2])
rd_at_site$Lon <- unlist(lons[,2])

df <- data.frame(
  lon = rd_at_site$Lon,
  lat = rd_at_site$Lat,
  RD = rd_at_site$RD
)

library(ggplot2)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)


# World map
world <- ne_countries(scale = "medium", returnclass = "sf")

# Get bounding box with a small padding
lon_range <- range(df$lat, na.rm = TRUE)
lat_range <- range(df$lon, na.rm = TRUE)

padding <- 0.2  # adjust as needed
xlim <- c(lon_range[1] - padding, lon_range[2] + padding)
ylim <- c(lat_range[1] - padding, lat_range[2] + padding)

# Plot
ggplot() +
  geom_sf(data = world, fill = "gray", color = "black") +
  geom_point(data = df, aes(x = lat, y = lon, color = RD), cex = 1) +
  scale_color_viridis_c(option = "inferno") +
  coord_sf(xlim = xlim, ylim = ylim, expand = FALSE) +
  theme_minimal() +
  labs(title = "RD across the WG", x = "Longitude", y = "Latitude", color = "RD") 
  


```


